Fast Bayesian Inference for Exoplanet Discovery in Radial Velocity Data

### 2014-11-13 Paper/ms.tex
Waffling about phase transitions

@@

The marginal likelihood was $\ln(\mathcal{Z}) \approx -618.04$ and the
information was $\mathcal{H} \approx 140.2$ nats. This compares favourably
to the marginal likelihood of $-640.1$ (for a 6-planet model)
found by \citet{fengji}, although it is unclear whether we used exactly the
same dataset.
Interestingly, the log-likelihood curve (Figure~\ref{fig:logl})
shows this problem has two phase transitions.

@@

\section{Conclusions}
In this paper we introduced a trans-dimensional MCMC approach to inferring
the number of planets $N$ in an exoplanetary system from radial velocity data.
The MCMC was implemented using the framework of \citet{rjobject} which
defines trans-dimensional birth and death moves, and does the sampling
with respect to a Nested Sampling target distribution, rather than directly
sampling the posterior. This approach allows us to compute the results in a
single run, which provides posterior samples and an estimate of the marginal
likelihood.

We applied the code to two well-studied RV datasets, $\nu$ Oph and Gliese 581.
In $\nu$ Oph, we found some evidence for a signal with a period of
75.7 $\pm$ 1.8 days: the posterior probability it exists, given our assumptions,
is 70\%. With the combined HIRES+HARPS dataset from Gliese 581, we found
evidence for a large number of ``planets'', although only four have well
determined periods, corresponding to the Gliese 581 b, c, d,
and e. Since our model does not include any possibility of stellar variability,
any such periodic signals will be attributed to planets. Including non-planetary
stellar variability is a crucial next step.

>>>>

\begin{figure}
\includegraphics[scale=0.45]{Figures/logl0.eps}
\caption{Log likelihood vs. enclosed prior mass (a standard output of Nested
Sampling algorithms) for the $\nu$ Oph analysis.
There are several phase transitions (concave-up regions) present. The one at
$\log(X) \approx -75$ separates models which contain the third signal from
models that do not. Without using Nested Sampling it would be very difficult
to mix between these two situations and calculate the posterior probability
for the existence of the third signal.
\label{fig:logl0}}
\end{figure}

Figure~\ref{fig:logl0} shows the relationship between the likelihood $L$
and the enclosed prior mass $X$ for the $\nu$ Oph analysis. These plots are
a standard output of Nested Sampling \citep{skilling} analyses, and provide
insights into the structure of the problem. Concave-up regions of this curve
indicate phase transitions which can cause severe problems for annealing-based
methods, and sometimes even for sampling the posterior distribution. In this
analysis, the models with a third signal exist to the left of the phase
transition at $\log(X) \approx -75$, and models without the signal exist to the
right of the phase transition. Mixing between these two phases is crucial for
accurately computing the posterior probability that the third signal exists.
However, if we had simply tried to sample the posterior distribution, it would
not mix between these two phases because the posterior is composed of a
large-volume low-likelihood ``slab'' and a low-volume high-likelihood ``spike''.
It is difficult for standard MCMC methods to escape the spike because of the
likelihood contrast, and it is also difficult to escape the slab and enter the
spike because of its small volume. Thus, even unimodal posterior distributions
can cause mixing problems analogous to multimodal situations if a phase
transition is present.

@@

The marginal likelihood was $\ln(\mathcal{Z}) \approx -618.04$ and the
information was $\mathcal{H} \approx 140.2$ nats. This compares favourably
to the marginal likelihood of $-640.1$ (for a 6-planet model)
found by \citet{fengji}, although it is unclear whether we used exactly the
same dataset.
Interestingly, the log-likelihood curve (Figure~\ref{fig:logl})
shows this problem has two phase transitions. While these do not affect the
posterior distribution (as they did for $\nu$ Oph), they would cause difficulties
if we tried to calculte the marginal likelihood using thermal annealing.

@@

\section{Conclusions}
In this paper we introduced a trans-dimensional MCMC approach to inferring
the number of planets $N$ in an exoplanetary system from radial velocity data.
The MCMC was implemented using the framework of \citet{rjobject} which
defines trans-dimensional birth and death moves, and does the sampling
with respect to a Nested Sampling target distribution, rather than directly
sampling the posterior. This approach allows us to compute the results in a
single run, which provides posterior samples and an estimate of the marginal
likelihood. By embedding the MCMC in a Nested Sampling algorithm, instead of
directly trying to sample the posterior distribution, we can overcome difficult
features in the problem, such as phase transitions.

We applied the code to two well-studied RV datasets, $\nu$ Oph and Gliese 581.
In $\nu$ Oph, we found some evidence for a signal with a period of
75.7 $\pm$ 1.8 days: the posterior probability it exists, given our assumptions,
is 70\%. The posterior distribution contains models with three signals and
models with two, however, these are separated by a phase transition. Therefore
mixing between the two situations would be very difficult if we simply tried
to sample the posterior distribution.

With the combined HIRES+HARPS dataset from Gliese 581, we found
evidence for a large number of ``planets'', although only four have well
determined periods, corresponding to the Gliese 581 b, c, d,
and e. Since our model does not include any possibility of stellar variability,
any such periodic signals will be attributed to ``planets''.
Including non-planetary stellar variability is a crucial next step.
